---
title: "Coral cover simulations"
author: "Robin Elahi"
date: "2017-12-14"
biblio-style: apsr
bibliography: ~/github/bib/master.bib
#csl: frontiers-in-ecology-and-the-environment.csl
endnote: no
fontfamily: mathpazo
fontsize: 11pt
geometry: margin=1in
output:
  pdf_document:
    fig_caption: no
    keep_tex: no
    latex_engine: pdflatex
  word_document: default
header-includes:
    - \usepackage{lineno}
    - \linenumbers  
---

```{r setup, include = FALSE, cache = FALSE}
knitr::opts_chunk$set(echo = FALSE, message = FALSE, warning = FALSE, 
                      fig.path = 'figs/', cache.path = 'cache/graphics-', 
                      fig.align = 'center', fig.width = 5, fig.height = 5, 
                      fig.show = 'hold', cache = TRUE, par = TRUE)
knitr::opts_knit$set(root.dir = "../")
```

```{r r_packages}
# tidyverse
library(dplyr); library(ggplot2); library(tidyr); library(readr)
theme_set(theme_bw(base_size = 12))
library(knitr)
```

## Introduction
Numerical simulations were based on modelled coral cover representing time-series analysis of a coral reef experiencing a variety of prescribed disturbance histories. 

### Stable
For sites exhibiting stable coral cover ($y_i$) in year $i$, we generated 30 random values about an intercept ($b$) with variation ($w_i$), for a thirty-year time-series as follows:

$$y_i = b + w_i$$
$$b = \mathcal{N}(40,5)~|~\mathcal{U}(5,40)$$
$$ \sigma_b = \mathcal{N}(0.1b, 0.5)$$
$$ w_i = \mathcal{N}(0, \sigma_b)$$
where $\sigma_b$ is the standard deviation about $b$.  The intercept $b$ is chosen from either a normal ($\mathcal{N}$) or uniform distribution ($\mathcal{U}$), determined by a fair coin flip. 

### Phase shift
For sites exhibiting a phase shift, a new intercept ($b_{new}$) and variation ($w_{i_{new}}$) was selected at a year chosen randomly between year 10 and year 30 ($x_{shift}$) of a thirty-year time-series as follows:

$$ y_i = \begin{cases} b + w_i & \text{if } x < x_{shift} \\ b_{new} + w_{i_{new}} & \text{if } x\ge x_{shift}  \end{cases} $$

$$ x_{shift} = \mathcal{U}(10,30) $$
$$ b_{new} = \mathcal{U}(1,0.5b)$$
$$ \sigma_{b_{new}} = \mathcal{N}(\sigma_{b}, 0.1)$$
$$ w_{i_{new}} = \mathcal{N}(0, \sigma_{b_{new}})$$
where $\sigma_{b_{new}}$ is the new standard deviation about $b_{new}$. The effect of the phase shift is to reduce the coral cover to a value chosen uniformly between 1 and half of the original $b$.   

### Linear trend
For sites exhibiting a linear trend in coral cover, a slope (m) was also selected randomly with a thirty-year time series determined by the equation for a line, as follows:

$$y_i = mx + b + w_i$$
$$b = \mathcal{N}(40,5)~|~\mathcal{U}(5,40)$$
$$ \sigma_b = \mathcal{N}(0.1b, 0.5)$$
$$ w_i = \mathcal{N}(0, \sigma_b)$$
$$ m = \mathcal{N}(-0.5, 0.25)$$
where $\sigma_b$ is the standard deviation about $b$.    

### Oscillations
Finally, we used a cosine curve (y = a*cos (bx + c) + w) to simulate oscillations over time, with an amplitude (a) and phase shift (c). 
